ExpressionView file format

### 2010-05-05 RExpressionView/inst/doc/ExpressionView.format.Rnw
EV file format vignette: better spacing.
EV format vignette: use a macro to typeset the XML tags.

\definecolor{green}{rgb}{0.8,1,0.8}
\definecolor{blue}{rgb}{0.8,0.8,1}
\lstset{basicstyle=\footnotesize\ttfamily,%
        keywordstyle=\footnotesize\ttfamily,%
        language=XML,frame=single,columns=fullflexible}
\lstnewenvironment{schema}{\lstset{backgroundcolor=\color{green}}}{}
\lstnewenvironment{XML}{\lstset{backgroundcolor=\color{blue}}}{}

@@

\begin{schema}
<xsd:element name="evf" type="evfType" />
\end{schema}
An EVF file contains a single \texttt{<evf>} tag. It has the following
parts:
\begin{schema}
<xsd:complexType name="evfType">
  <xsd:sequence>
    <xsd:element name="summary" type="SummaryType" />
    <xsd:element name="experimentdata" type="ExperimentDataType" />
    <xsd:element name="genes" type="GenesType" />
    <xsd:element name="samples" type="SamplesType" />
    <xsd:element name="modules" type="ModulesType" />
    <xsd:element name="data" type="xsd:base64Binary" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}

The \texttt{<summary>} tag contains information about the data set, such
as the number of genes, samples, modules, etc. \texttt{<experimentdata>}
is for experiment meta-data, i.e. the lab where it was performed, 
the abstract of the related publication and possibly
more. \texttt{<genes>} and \texttt{<samples>} have the gene and sample
meta data. \texttt{<modules>} defines the biclusters. Finally
\texttt{<data>} contains the expression values themselves. Let us see
all of these tags in detail now.

@@

This is the type of the \texttt{<summary>} tag:
\begin{schema}  
<xsd:complexType name="SummaryType">
  <xsd:sequence>
    <xsd:element name="description" minOccurs="0" type="xsd:string" />
    <xsd:element name="version" type="xsd:string" />
    <xsd:element name="dataorigin" minOccurs="0" type="xsd:string" />
    <xsd:element name="xaxislabel" minOccurs="0" type="xsd:string" />
    <xsd:element name="yaxislabel" minOccurs="0" type="xsd:string" />
    <xsd:element name="nmodules" type="xsd:nonNegativeInteger" />
    <xsd:element name="ngenes" type="xsd:positiveInteger" />
    <xsd:element name="nsamples" type="xsd:positiveInteger" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}
The \texttt{<summary>} tag optionally contains the description of the
data file (\texttt{<description>} tag), this is displayed above the main
gene expression window in ExpressionView. The \texttt{<version>} tag
must be \texttt{1.1} for files for the format we are discussing here. The
\texttt{<dataorigin>} tag is optional and has the value `\texttt{eisa}' for
modules generated with the ISA algorithm~\cite{bergmann03} and
exported using the ExpressionView R package. \texttt{<xaxislabel>} and
\texttt{<yaxislabel>} are optional axis labels for the gene expression
plot. The last three tags (\texttt{<nmodules>}, \texttt{<ngenes>} and
\texttt{<nsamples>}) are required and give the number of modules,
number of genes and number of samples in the data set.

A sample EVF file header, together with the \texttt{<summary>} tag looks
like this:
\begin{XML}
<?xml version="1.0" encoding="UTF-8"?>
<evf>
  <summary>
    <description>ExpressionView data file</description>
    <version>1.1</version>
    <dataorigin>eisa</dataorigin>
    <nmodules>8</nmodules>
    <ngenes>3522</ngenes>
    <nsamples>128</nsamples>
  </summary>
\end{XML}

@@

The \texttt{<experimentdata>} tag is next, this contains the
experiment meta-data. It has the following parts:
\begin{schema}
<xsd:complexType name="ExperimentDataType">
  <xsd:all>
    <xsd:element name="title" minOccurs="0" type="xsd:string" />
    <xsd:element name="name" minOccurs="0" type="xsd:string" />
    <xsd:element name="lab" minOccurs="0" type="xsd:string" />
    <xsd:element name="abstract" minOccurs="0" type="xsd:string" />
    <xsd:element name="url" minOccurs="0" type="xsd:anyURI" />
    <xsd:element name="annotation" minOccurs="0" type="xsd:string" />
    <xsd:element name="organism" minOccurs="0" type="xsd:string" />
  </xsd:all>
</xsd:complexType>
\end{schema}
All fields are optional here: \texttt{<title>}, \texttt{<name>},
\texttt{<lab>}, \texttt{<abstract>}, \texttt{<url>},
\texttt{<annotation>}, \texttt{<organism>}. They are collected and
shown in the \emph{Experiment} tab in the interactive ExpressionView
viewer. The \texttt{<annotation>} tag should contain the name of the
chip on which the experiment was performed. If the \texttt{<organism>}
tag contains `\texttt{Homo sapiens}', then ExpressionView links genes with
the Gene Cards homepage, for other organism is uses Entrez Gene. The
tags within \texttt{experimentdata} can be specified in arbitrary
order.

Here is an example for the \texttt{<experimentdata>} tag:
\begin{XML}
<experimentdata>
  <title>Gene expression profile of adult T-cell acute
    lymphocytic leukemia identifies distinct subsets of
    patients with different response to therapy and
    survival.
  </title> 
  <name>Chiaretti et al.</name>
  <lab>Department of Medical Oncology, ...</lab>
  <abstract>Gene expression profiles were examined ...</abstact>
  <url>http://...</url>
  <annotation>hgu95av2</annotation>
  <organism>Homo sapiens</organism>
</experimentdata>
\end{XML}

@@

The next tag within \texttt{<evf>} is \texttt{<genes>}, its type os
defined as
\begin{schema}  
<xsd:complexType name="GenesType">
  <xsd:sequence>
    <xsd:element name="genetags" type="GeneTagsType" />
    <xsd:element name="gene" maxOccurs="unbounded" type="GeneType" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}
It has two main parts, a \texttt{<genetags>} tag first, and then a
number of \texttt{<gene>} tags, one for each gene in the data set.

@@

The \texttt{<genetags>} tag has one required and many optional
subtags:
\begin{schema}  
<xsd:complexType name="GeneTagsType">
  <xsd:sequence>
    <xsd:element name="id" type="xsd:string" />
    <xsd:element name="name" minOccurs="0" type="xsd:string" />
    <xsd:element name="symbol" minOccurs="0" type="xsd:string" />
    <xsd:element name="entrezid" minOccurs="0" type="xsd:string" />
    <xsd:element name="x" minOccurs="0" maxOccurs="unbounded"
                 type="ExtraType" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}
The \texttt{<id>} must be an integer number, that starts with one and
goes up to the number of genes. It is used as an id that is refered
from other tags in the document. \texttt{<name>} is typically the name
of the probe-set, but it can be used for other purposes as well,
\texttt{<symbol>} is typically the canonical gene
symbol. \texttt{<entrezid>} is the Entrez gene id. If a given
probeset does not map to a known gene, then \texttt{<symbol>} and
\texttt{<entrezid>} can be set to \texttt{NA}. Finally, the gene tags
might contain any number of \texttt{<x>} tags, each defining an
additional column in the \emph{Genes} tab in ExpressionView. These
tags must have a \texttt{name} attribute, which is refered by the
tags of the individual genes. Here is an example gene tags section:
\begin{XML}
<genetags>
  <id>#</id>
  <name>Name</name>
  <symbol>Symbol</symbol>
  <entrezid>EntrezID</entrezid>
  <x name="chr">Chromosome</x>
</genetags>
\end{XML}
We define an extra column for the gene table, this will give the
chromosome on which the gene is located.

After the gene tags, we have a typically large number of genes in the
file. The definition of the \texttt{<gene>} tag:
\begin{schema}
<xsd:complexType name="GeneType">
  <xsd:sequence>
    <xsd:element name="id" type="xsd:string" />
    <xsd:element name="name" minOccurs="0" type="xsd:string" />
    <xsd:element name="symbol" minOccurs="0" type="xsd:string" />
    <xsd:element name="entrezid" minOccurs="0" type="xsd:string" />
    <xsd:element name="x" minOccurs="0" maxOccurs="unbounded"
                 type="ExtraType"/>
  </xsd:sequence>
</xsd:complexType>
\end{schema}
The \texttt{<gene>} tags must have exactly the same subtags as the
ones given in the gene tags section. For example, assuming the
\texttt{<genetags>} example above we can have:
\begin{XML}
<gene>
  <id>1</id>
  <name>33500_i_at</name>
  <symbol>NA</symbol>
  <entrezid>NA</entrezid>
  <x name="chr">NA</x>
</gene>
<gene>
  <id>2</id>
  <name>40990_at</name>
  <symbol>TSPAN5</symbol>
  <entrezid>10098</entrezid>
  <x name="chr">4</x>
</gene>
...
\end{XML}
Observe, how the extra tag for the chromosomes is referenced here.

@@

The samples and their associated data have a format similar to the
genes:
\begin{schema}  
<xsd:complexType name="SamplesType">
  <xsd:sequence>
    <xsd:element name="sampletags" type="SampleTagsType" />
    <xsd:element name="sample" maxOccurs="unbounded" 
                 type="SampleType" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}
The \texttt{<samples>} tag has two parts, the first defines the sample
meta data entries and the second part contains one \texttt{<sample>}
tag for each sample in the data set.

\begin{schema}  
<xsd:complexType name="SampleTagsType">
  <xsd:sequence>
    <xsd:element name="id" type="xsd:string" />
    <xsd:element name="name" minOccurs="0" type="xsd:string" />
    <xsd:element name="x" minOccurs="0" maxOccurs="unbounded"
                 type="ExtraType" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}
The \texttt{<id>} and \texttt{<name>} sample tags are
required. \texttt{<id>} contains numeric ids, from one, up to the
number of samples in the data set. These are referenced by other tags,
e.g. the ones that define the modules. \texttt{<name>} is an arbitrary
string, it is typically the sample name in the R ExpressionSet object,
that was used to find the biclusters. Similarly to the genes, any
number of additional \texttt{<x>} tags can be added, each with a
unique \texttt{name} attribute, to define additional information about
the samples.

Here is an example \texttt{<sampletags>} tag:
\begin{XML}
<sampletags>
  <id>#</id>
  <name>Name</name>
  <x name="diagnosis"> Date of diagnosis</x>
  <x name="sex"> Gender of the patient</x>
  <x name="age"> Age of the patient at entry</x>
  <x name="BT"> does the patient have B-cell or T-cell ALL</x>
</sampletags>
\end{XML}
We defined four extra tags for the samples in this example.

Then the samples follow:
\begin{schema}  
<xsd:complexType name="SampleType">
  <xsd:sequence>
    <xsd:element name="id" type="xsd:string" />
    <xsd:element name="name" minOccurs="0" type="xsd:string" />
    <xsd:element name="x" minOccurs="0" maxOccurs="unbounded"
                 type="ExtraType" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}
Here is an example, that corresponds to the \texttt{<sampletags>}
entry given above:
\begin{XML}
<sample>
  <id>1</id>
  <name>01010</name>
  <x name="diagnosis">3/29/2000</x>
  <x name="sex">2</x>
  <x name="age">19</x>
  <x name="BT">3</x>
</sample>
<sample>
  <id>2</id>
  <name>04010</name>
  <x name="diagnosis">10/30/1997</x>
  <x name="sex">1</x>
  <x name="age">18</x>
  <x name="BT">2</x>
</sample>
...
\end{XML}

@@

Again, the \texttt{<modules>} tag has a syntax that is similar to the
syntax of the genes and the samples:
\begin{schema}
<xsd:complexType name="ModulesType">
  <xsd:sequence>
    <xsd:element name="moduletags" type="ModuleTagsType" />
    <xsd:element name="gotags" minOccurs="0" type="GoTagsType" />
    <xsd:element name="keggtags" minOccurs="0" type="KeggTagsType" />
    <xsd:element name="module" maxOccurs="unbounded" 
                 type="ModuleType" />
  </xsd:sequence>      
</xsd:complexType>
\end{schema}
There are two required and two optional parts. The first part is
\texttt{<moduletags>}, this defines the meta-data associated with the
modules. The \texttt{<gotags>} and \texttt{<keggtags>} tags are
optional, they are only present if the file contains Gene Ontology
and/or KEGG pathway enrichment calculation results for the
biclusters. Finally, there is a list of \texttt{<module>} tags, one
for each bicluster.

The type of \texttt{<moduletags>}:
\begin{schema}  
<xsd:complexType name="ModuleTagsType">
  <xsd:sequence>
    <xsd:element name="id" type="xsd:string" />
    <xsd:element name="name" minOccurs="0" type="xsd:string" />
    <xsd:element name="iterations" minOccurs="0" type="xsd:string" />
    <xsd:element name="oscillation" minOccurs="0" type="xsd:string" />
    <xsd:element name="thr_row" minOccurs="0" type="xsd:string" />
    <xsd:element name="thr_col" minOccurs="0" type="xsd:string" />
    <xsd:element name="freq" minOccurs="0" type="xsd:string" />
    <xsd:element name="rob" minOccurs="0" type="xsd:string" />
    <xsd:element name="rob_limit" minOccurs="0" type="xsd:string" />
    <xsd:element name="x" minOccurs="0" maxOccurs="unbounded"
		 type="ExtraType" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}
There is only one tag required in \texttt{<moduletags>}, the
\texttt{<id>}, which is a numeric id, going from 1 to the number of
modules. The rest is optional, and it is also possible to create any
kind of extra tags with the \texttt{<x>} tag. \texttt{<name>} is an
arbitrary character string. The other optional tags are
mainly defined to modules coming from the ISA algorithm:
\texttt{<iterations>} gives the number of ISA iterations needed to
find the module, \texttt{<oscillation>} specifies the oscillation
cycle length for oscillating modules. It is mostly zero, meaning that
the module does not oscillate. \texttt{<thr\_row>} is the ISA gene
threshold, \texttt{<thr\_col>} is the ISA condition
threshold. \texttt{<freq>} is the number of ISA seeds that converged
to the module, \texttt{<rob>} is its robustness
score. \texttt{<rob\_limit>} is the robustness threshold that was used
to filter the module.

Here is an example \texttt{<moduletags>} tag:
\begin{XML}
<moduletags>
  <id>#</id>
  <name>Name</name>
  <iterations>Iterations</iterations>
  <oscillation>Oscillation cycle</oscillation>
  <thr_row>Gene threshold</thr_row>
  <thr_col>Column threshold</thr_col>
  <freq>Frequency</freq>
  <rob>Robustness</rob>
  <rob_limit>Robustness limit</rob_limit>
</moduletags>
\end{XML}

The optional \texttt{<gotags>} and \texttt{<keggtags>} tags are
discussed in Section~\ref{sec:gokegg}, let us now see how the data for a
single module looks like:
\begin{schema}
<xsd:complexType name="ModuleType">
  <xsd:sequence>
    <xsd:element name="id" type="xsd:string" />
    <xsd:element name="name" minOccurs="0" type="xsd:string" />
    <xsd:element name="iterations" minOccurs="0" type="xsd:string" />
    <xsd:element name="oscillation" minOccurs="0" type="xsd:string" />
    <xsd:element name="thr_row" minOccurs="0" type="xsd:string" />
    <xsd:element name="thr_col" minOccurs="0" type="xsd:string" />
    <xsd:element name="freq" minOccurs="0" type="xsd:string" />
    <xsd:element name="rob" minOccurs="0" type="xsd:string" />
    <xsd:element name="rob_limit" minOccurs="0" type="xsd:string" />
    <xsd:element name="x" minOccurs="0" maxOccurs="unbounded"
		 type="ExtraType" />
    <xsd:element name="containedgenes" type="integerList" />
    <xsd:element name="genescores" type="doubleList" />
    <xsd:element name="containedsamples" type="integerList" />
    <xsd:element name="samplescores" type="doubleList" />
    <xsd:element name="intersectingmodules" type="integerList" />
    <xsd:element name="gos" minOccurs="0" type="GosType" />
    <xsd:element name="keggs" minOccurs="0" type="KeggsType" />
  </xsd:sequence>
</xsd:complexType>  
\end{schema}
The first couple of fields refer to the ones defined in the
\texttt{<moduletags>} section above. The others are:
\texttt{<containedgenes>}, a space separated list of gene ids, for the
genes that are included in the module; \texttt{<genescores>}, the gene
scores of these genes, in the order of the list in
\texttt{<containedgenes>}; \texttt{<containedsamples>}, the ids of the
samples in the module; \texttt{<samplescores>}, the scores for these
samples; \texttt{<intersectingmodules>}, the ids of the modules that
have an overlap with the given module. \texttt{<gos>} and
\texttt{<keggs>} are optional and describe the Gene Ontology and KEGG
pathway enrichment of the given module, see their details in Section~\ref{sec:gokegg}.

Here is an example \texttt{<module>} tag:
\begin{XML}
<module>
  <id>1</id>
  <name>module 1</name>
  <iterations>22</iterations>
  <oscillation>0</oscillation>
  <thr_row>2.7</thr_row>
  <thr_col>1.4</thr_col>
  <freq>1</freq>
  <rob>21.98</rob>
  <rob_limit>21.98</rob_limit>
  <containedgenes>214 215 216 217 218 219 220 221 222
  </containedgenes>
  <genescores>-0.94 -0.88 0.74 -0.76 -1.00 -0.84 -0.74 -0.76 -0.85
  </genescores>
  <containedsamples>63 64 65 54 66</containedsamples>
  <samplescores>-0.62 -1.00 -0.77 -0.40 -0.28</samplescores>
  <intersectingmodules>7</intersectingmodules>
  <gos>
    ... see below ...
  </gos>
  <keggs>
    ... see below ...
  </keggs>
</module>
\end{XML}

>>>>

\definecolor{green}{rgb}{0.8,1,0.8}
\definecolor{blue}{rgb}{0.8,0.8,1}
\lstset{basicstyle=\footnotesize\ttfamily,%
        keywordstyle=\footnotesize\ttfamily,%
        language=XML,frame=single,columns=fullflexible,%
        aboveskip=8pt,belowskip=-5pt}
\lstnewenvironment{schema}{\lstset{backgroundcolor=\color{green}}}{}
\lstnewenvironment{example}{\lstset{backgroundcolor=\color{blue}}}{}

\newcommand{\xmltag}[1]{%
  \csname a:xmltag\endcsname\texttt{<#1>}\csname b:xmltag\endcsname}

@@

\begin{schema}
<xsd:element name="evf" type="evfType" />
\end{schema}
An EVF file contains a single \xmltag{evf} tag. It has the following
parts:
\begin{schema}
<xsd:complexType name="evfType">
  <xsd:sequence>
    <xsd:element name="summary" type="SummaryType" />
    <xsd:element name="experimentdata" type="ExperimentDataType" />
    <xsd:element name="genes" type="GenesType" />
    <xsd:element name="samples" type="SamplesType" />
    <xsd:element name="modules" type="ModulesType" />
    <xsd:element name="data" type="xsd:base64Binary" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}

The \xmltag{summary} tag contains information about the data set, such
as the number of genes, samples, modules, etc. \xmltag{experimentdata}
is for experiment meta-data, i.e. the lab where it was performed, 
the abstract of the related publication and possibly
more. \xmltag{genes} and \xmltag{samples} have the gene and sample
meta data. \xmltag{modules} defines the biclusters. Finally
\xmltag{data} contains the expression values themselves. Let us see
all of these tags in detail now.

@@

This is the type of the \xmltag{summary} tag:
\begin{schema}  
<xsd:complexType name="SummaryType">
  <xsd:sequence>
    <xsd:element name="description" minOccurs="0" type="xsd:string" />
    <xsd:element name="version" type="xsd:string" />
    <xsd:element name="dataorigin" minOccurs="0" type="xsd:string" />
    <xsd:element name="xaxislabel" minOccurs="0" type="xsd:string" />
    <xsd:element name="yaxislabel" minOccurs="0" type="xsd:string" />
    <xsd:element name="nmodules" type="xsd:nonNegativeInteger" />
    <xsd:element name="ngenes" type="xsd:positiveInteger" />
    <xsd:element name="nsamples" type="xsd:positiveInteger" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}
The \xmltag{summary} tag optionally contains the description of the
data file (\xmltag{description} tag), this is displayed above the main
gene expression window in ExpressionView. The \xmltag{version} tag
must be \texttt{1.1} for files for the format we are discussing here. The
\xmltag{dataorigin} tag is optional and has the value `\texttt{eisa}' for
modules generated with the ISA algorithm~\cite{bergmann03} and
exported using the ExpressionView R package. \xmltag{xaxislabel} and
\xmltag{yaxislabel} are optional axis labels for the gene expression
plot. The last three tags (\xmltag{nmodules}, \xmltag{ngenes} and
\xmltag{nsamples}) are required and give the number of modules,
number of genes and number of samples in the data set.

A sample EVF file header, together with the \xmltag{summary} tag looks
like this:
\begin{example}
<?xml version="1.0" encoding="UTF-8"?>
<evf>
  <summary>
    <description>ExpressionView data file</description>
    <version>1.1</version>
    <dataorigin>eisa</dataorigin>
    <nmodules>8</nmodules>
    <ngenes>3522</ngenes>
    <nsamples>128</nsamples>
  </summary>
\end{example}

@@

The \xmltag{experimentdata} tag is next, this contains the
experiment meta-data. It has the following parts:
\begin{schema}
<xsd:complexType name="ExperimentDataType">
  <xsd:all>
    <xsd:element name="title" minOccurs="0" type="xsd:string" />
    <xsd:element name="name" minOccurs="0" type="xsd:string" />
    <xsd:element name="lab" minOccurs="0" type="xsd:string" />
    <xsd:element name="abstract" minOccurs="0" type="xsd:string" />
    <xsd:element name="url" minOccurs="0" type="xsd:anyURI" />
    <xsd:element name="annotation" minOccurs="0" type="xsd:string" />
    <xsd:element name="organism" minOccurs="0" type="xsd:string" />
  </xsd:all>
</xsd:complexType>
\end{schema}
All fields are optional here: \xmltag{title}, \xmltag{name},
\xmltag{lab}, \xmltag{abstract}, \xmltag{url},
\xmltag{annotation}, \xmltag{organism}. They are collected and
shown in the \emph{Experiment} tab in the interactive ExpressionView
viewer. The \xmltag{annotation} tag should contain the name of the
chip on which the experiment was performed. If the \xmltag{organism}
tag contains `\texttt{Homo sapiens}', then ExpressionView links genes with
the Gene Cards homepage, for other organism is uses Entrez Gene. The
tags within \texttt{experimentdata} can be specified in arbitrary
order.

Here is an example for the \xmltag{experimentdata} tag:
\begin{example}
<experimentdata>
  <title>Gene expression profile of adult T-cell acute
    lymphocytic leukemia identifies distinct subsets of
    patients with different response to therapy and
    survival.
  </title> 
  <name>Chiaretti et al.</name>
  <lab>Department of Medical Oncology, ...</lab>
  <abstract>Gene expression profiles were examined ...</abstact>
  <url>http://...</url>
  <annotation>hgu95av2</annotation>
  <organism>Homo sapiens</organism>
</experimentdata>
\end{example}

@@

The next tag within \xmltag{evf} is \xmltag{genes}, its type os
defined as
\begin{schema}  
<xsd:complexType name="GenesType">
  <xsd:sequence>
    <xsd:element name="genetags" type="GeneTagsType" />
    <xsd:element name="gene" maxOccurs="unbounded" type="GeneType" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}
It has two main parts, a \xmltag{genetags} tag first, and then a
number of \xmltag{gene} tags, one for each gene in the data set.

@@

The \xmltag{genetags} tag has one required and many optional
subtags:
\begin{schema}  
<xsd:complexType name="GeneTagsType">
  <xsd:sequence>
    <xsd:element name="id" type="xsd:string" />
    <xsd:element name="name" minOccurs="0" type="xsd:string" />
    <xsd:element name="symbol" minOccurs="0" type="xsd:string" />
    <xsd:element name="entrezid" minOccurs="0" type="xsd:string" />
    <xsd:element name="x" minOccurs="0" maxOccurs="unbounded"
                 type="ExtraType" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}
The \xmltag{id} must be an integer number, that starts with one and
goes up to the number of genes. It is used as an id that is refered
from other tags in the document. \xmltag{name} is typically the name
of the probe-set, but it can be used for other purposes as well,
\xmltag{symbol} is typically the canonical gene
symbol. \xmltag{entrezid} is the Entrez gene id. If a given
probeset does not map to a known gene, then \xmltag{symbol} and
\xmltag{entrezid} can be set to \texttt{NA}. Finally, the gene tags
might contain any number of \xmltag{x} tags, each defining an
additional column in the \emph{Genes} tab in ExpressionView. These
tags must have a \texttt{name} attribute, which is refered by the
tags of the individual genes. Here is an example gene tags section:
\begin{example}
<genetags>
  <id>#</id>
  <name>Name</name>
  <symbol>Symbol</symbol>
  <entrezid>EntrezID</entrezid>
  <x name="chr">Chromosome</x>
</genetags>
\end{example}
We define an extra column for the gene table, this will give the
chromosome on which the gene is located.

After the gene tags, we have a typically large number of genes in the
file. The definition of the \xmltag{gene} tag:
\begin{schema}
<xsd:complexType name="GeneType">
  <xsd:sequence>
    <xsd:element name="id" type="xsd:string" />
    <xsd:element name="name" minOccurs="0" type="xsd:string" />
    <xsd:element name="symbol" minOccurs="0" type="xsd:string" />
    <xsd:element name="entrezid" minOccurs="0" type="xsd:string" />
    <xsd:element name="x" minOccurs="0" maxOccurs="unbounded"
                 type="ExtraType"/>
  </xsd:sequence>
</xsd:complexType>
\end{schema}
The \xmltag{gene} tags must have exactly the same subtags as the
ones given in the gene tags section. For example, assuming the
\xmltag{genetags} example above we can have:
\begin{example}
<gene>
  <id>1</id>
  <name>33500_i_at</name>
  <symbol>NA</symbol>
  <entrezid>NA</entrezid>
  <x name="chr">NA</x>
</gene>
<gene>
  <id>2</id>
  <name>40990_at</name>
  <symbol>TSPAN5</symbol>
  <entrezid>10098</entrezid>
  <x name="chr">4</x>
</gene>
...
\end{example}
Observe, how the extra tag for the chromosomes is referenced here.

@@

The samples and their associated data have a format similar to the
genes:
\begin{schema}  
<xsd:complexType name="SamplesType">
  <xsd:sequence>
    <xsd:element name="sampletags" type="SampleTagsType" />
    <xsd:element name="sample" maxOccurs="unbounded" 
                 type="SampleType" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}
The \xmltag{samples} tag has two parts, the first defines the sample
meta data entries and the second part contains one \xmltag{sample}
tag for each sample in the data set.

\begin{schema}  
<xsd:complexType name="SampleTagsType">
  <xsd:sequence>
    <xsd:element name="id" type="xsd:string" />
    <xsd:element name="name" minOccurs="0" type="xsd:string" />
    <xsd:element name="x" minOccurs="0" maxOccurs="unbounded"
                 type="ExtraType" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}
The \xmltag{id} and \xmltag{name} sample tags are
required. \xmltag{id} contains numeric ids, from one, up to the
number of samples in the data set. These are referenced by other tags,
e.g. the ones that define the modules. \xmltag{name} is an arbitrary
string, it is typically the sample name in the R ExpressionSet object,
that was used to find the biclusters. Similarly to the genes, any
number of additional \xmltag{x} tags can be added, each with a
unique \texttt{name} attribute, to define additional information about
the samples.

Here is an example \xmltag{sampletags} tag:
\begin{example}
<sampletags>
  <id>#</id>
  <name>Name</name>
  <x name="diagnosis"> Date of diagnosis</x>
  <x name="sex"> Gender of the patient</x>
  <x name="age"> Age of the patient at entry</x>
  <x name="BT"> does the patient have B-cell or T-cell ALL</x>
</sampletags>
\end{example}
We defined four extra tags for the samples in this example.

Then the samples follow:
\begin{schema}  
<xsd:complexType name="SampleType">
  <xsd:sequence>
    <xsd:element name="id" type="xsd:string" />
    <xsd:element name="name" minOccurs="0" type="xsd:string" />
    <xsd:element name="x" minOccurs="0" maxOccurs="unbounded"
                 type="ExtraType" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}
Here is an example, that corresponds to the \xmltag{sampletags}
entry given above:
\begin{example}
<sample>
  <id>1</id>
  <name>01010</name>
  <x name="diagnosis">3/29/2000</x>
  <x name="sex">2</x>
  <x name="age">19</x>
  <x name="BT">3</x>
</sample>
<sample>
  <id>2</id>
  <name>04010</name>
  <x name="diagnosis">10/30/1997</x>
  <x name="sex">1</x>
  <x name="age">18</x>
  <x name="BT">2</x>
</sample>
...
\end{example}

@@

Again, the \xmltag{modules} tag has a syntax that is similar to the
syntax of the genes and the samples:
\begin{schema}
<xsd:complexType name="ModulesType">
  <xsd:sequence>
    <xsd:element name="moduletags" type="ModuleTagsType" />
    <xsd:element name="gotags" minOccurs="0" type="GoTagsType" />
    <xsd:element name="keggtags" minOccurs="0" type="KeggTagsType" />
    <xsd:element name="module" maxOccurs="unbounded" 
                 type="ModuleType" />
  </xsd:sequence>      
</xsd:complexType>
\end{schema}
There are two required and two optional parts. The first part is
\xmltag{moduletags}, this defines the meta-data associated with the
modules. The \xmltag{gotags} and \xmltag{keggtags} tags are
optional, they are only present if the file contains Gene Ontology
and/or KEGG pathway enrichment calculation results for the
biclusters. Finally, there is a list of \xmltag{module} tags, one
for each bicluster.

The type of \xmltag{moduletags}:
\begin{schema}  
<xsd:complexType name="ModuleTagsType">
  <xsd:sequence>
    <xsd:element name="id" type="xsd:string" />
    <xsd:element name="name" minOccurs="0" type="xsd:string" />
    <xsd:element name="iterations" minOccurs="0" type="xsd:string" />
    <xsd:element name="oscillation" minOccurs="0" type="xsd:string" />
    <xsd:element name="thr_row" minOccurs="0" type="xsd:string" />
    <xsd:element name="thr_col" minOccurs="0" type="xsd:string" />
    <xsd:element name="freq" minOccurs="0" type="xsd:string" />
    <xsd:element name="rob" minOccurs="0" type="xsd:string" />
    <xsd:element name="rob_limit" minOccurs="0" type="xsd:string" />
    <xsd:element name="x" minOccurs="0" maxOccurs="unbounded"
                 type="ExtraType" />
  </xsd:sequence>
</xsd:complexType>
\end{schema}

There is only one tag required in \xmltag{moduletags}, the
\xmltag{id}, which is a numeric id, going from 1 to the number of
modules. The rest is optional, and it is also possible to create any
kind of extra tags with the \xmltag{x} tag. \xmltag{name} is an
arbitrary character string. The other optional tags are
mainly defined to modules coming from the ISA algorithm:
\xmltag{iterations} gives the number of ISA iterations needed to
find the module, \xmltag{oscillation} specifies the oscillation
cycle length for oscillating modules. It is mostly zero, meaning that
the module does not oscillate. \xmltag{thr\_row} is the ISA gene
threshold, \xmltag{thr\_col} is the ISA condition
threshold. \xmltag{freq} is the number of ISA seeds that converged
to the module, \xmltag{rob} is its robustness
score. \xmltag{rob\_limit} is the robustness threshold that was used
to filter the module.

Here is an example \xmltag{moduletags} tag:
\begin{example}
<moduletags>
  <id>#</id>
  <name>Name</name>
  <iterations>Iterations</iterations>
  <oscillation>Oscillation cycle</oscillation>
  <thr_row>Gene threshold</thr_row>
  <thr_col>Column threshold</thr_col>
  <freq>Frequency</freq>
  <rob>Robustness</rob>
  <rob_limit>Robustness limit</rob_limit>
</moduletags>
\end{example}

The optional \xmltag{gotags} and \xmltag{keggtags} tags are
discussed in Section~\ref{sec:gokegg}, let us now see how the data for a
single module looks like:
\begin{schema}
<xsd:complexType name="ModuleType">
  <xsd:sequence>
    <xsd:element name="id" type="xsd:string" />
    <xsd:element name="name" minOccurs="0" type="xsd:string" />
    <xsd:element name="iterations" minOccurs="0" type="xsd:string" />
    <xsd:element name="oscillation" minOccurs="0" type="xsd:string" />
    <xsd:element name="thr_row" minOccurs="0" type="xsd:string" />
    <xsd:element name="thr_col" minOccurs="0" type="xsd:string" />
    <xsd:element name="freq" minOccurs="0" type="xsd:string" />
    <xsd:element name="rob" minOccurs="0" type="xsd:string" />
    <xsd:element name="rob_limit" minOccurs="0" type="xsd:string" />
    <xsd:element name="x" minOccurs="0" maxOccurs="unbounded"
		 type="ExtraType" />
    <xsd:element name="containedgenes" type="integerList" />
    <xsd:element name="genescores" type="doubleList" />
    <xsd:element name="containedsamples" type="integerList" />
    <xsd:element name="samplescores" type="doubleList" />
    <xsd:element name="intersectingmodules" type="integerList" />
    <xsd:element name="gos" minOccurs="0" type="GosType" />
    <xsd:element name="keggs" minOccurs="0" type="KeggsType" />
  </xsd:sequence>
</xsd:complexType>  
\end{schema}
The first couple of fields refer to the ones defined in the
\xmltag{moduletags} section above. The others are:
\xmltag{containedgenes}, a space separated list of gene ids, for the
genes that are included in the module; \xmltag{genescores}, the gene
scores of these genes, in the order of the list in
\xmltag{containedgenes}; \xmltag{containedsamples}, the ids of the
samples in the module; \xmltag{samplescores}, the scores for these
samples; \xmltag{intersectingmodules}, the ids of the modules that
have an overlap with the given module. \xmltag{gos} and
\xmltag{keggs} are optional and describe the Gene Ontology and KEGG
pathway enrichment of the given module, see their details in Section~\ref{sec:gokegg}.

Here is an example \xmltag{module} tag:
\begin{example}
<module>
  <id>1</id>
  <name>module 1</name>
  <iterations>22</iterations>
  <oscillation>0</oscillation>
  <thr_row>2.7</thr_row>
  <thr_col>1.4</thr_col>
  <freq>1</freq>
  <rob>21.98</rob>
  <rob_limit>21.98</rob_limit>
  <containedgenes>214 215 216 217 218 219 220 221 222
  </containedgenes>
  <genescores>-0.94 -0.88 0.74 -0.76 -1.00 -0.84 -0.74 -0.76 -0.85
  </genescores>
  <containedsamples>63 64 65 54 66</containedsamples>
  <samplescores>-0.62 -1.00 -0.77 -0.40 -0.28</samplescores>
  <intersectingmodules>7</intersectingmodules>
  <gos>
    ... see below ...
  </gos>
  <keggs>
    ... see below ...
  </keggs>
</module>
\end{example}

We refered to the integer list and double list data types above, now
we define them, this is the integer list:
\begin{schema}  
<xsd:simpleType name="integerList">
  <xsd:list itemType="xsd:positiveInteger"/>
</xsd:simpleType>
\end{schema}

and similarly for the double list:
